---
title       : Getting Data (Part 2)
subtitle    : 
author      : Jeffrey Leek, Assistant Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn   # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 70)
opts_chunk$set(message = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache=F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```


## Interacting more directly with files

* file - open a connection to a text file
* url - open a connection to a url
* gzfile - open a connection to a .gz file
* bzfile - open a connection to a .bz2 file
* _?connections_ for more information
* <redtext>Remember to close connections </redtext>


---

## readLines() - local file

* readLines - a function to read lines of text from a connection
* Important parameters: _con_, _n_, _encoding_

```{r cachedChunk, echo=FALSE}
fileUrl <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.csv?accessType=DOWNLOAD"
download.file(fileUrl,destfile="./data/cameras.csv",method="curl")
```


```{r, dependson="cachedChunk"}
con <- file("./data/cameras.csv","r")
cameraData <- read.csv(con)
close(con)
head(cameraData)
```


---

## readLines() - from the web


```{r}

con <- url("http://simplystatistics.org","r")
simplyStats <- readLines(con)
close(con)
head(simplyStats)

```

---

## Reading JSON files {RJSONIO}


```{r}
library(RJSONIO)
fileUrl <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.json?accessType=DOWNLOAD"
download.file(fileUrl,destfile="./data/camera.json",method="curl")
con = file("./data/camera.json")
jsonCamera = fromJSON(con)
close(con)
head(jsonCamera)
```


---

## Writing data - write.table()

* The opposite of read.table
* Important parameters: _x_, _file_, _quote_, _sep_, _row.names_, _col.names_

```{r}
cameraData <- read.csv("./data/cameras.csv")
tmpData <- cameraData[,-1]
write.table(tmpData,file="./data/camerasModified.csv",sep=",")
cameraData2 <- read.csv("./data/camerasModified.csv")
head(cameraData2)
```


--- 

## Writing data - save(), save.image()


* save is used to save R objects
* Important parameters: _list of objects_, _file_
* save.image saves everything in your working directory

```{r}
cameraData <- read.csv("./data/cameras.csv")
tmpData <- cameraData[,-1]
save(tmpData,cameraData,file="./data/cameras.rda")

```


---

## Reading saved data - load()

* Opposite of save()
* Important parameters: _file_

```{r}
# Remove everything from the workspace
rm(list=ls())
ls()

# Load data 
load("./data/cameras.rda")
ls()

```


---

## paste() and paste0()

* These functions are for pasting character strings together. 
* Important parameters: _list of text strings_, _sep_
* paste0() is the same as paste but with _sep=""_
* Great for looping over files
* See also [file.path](http://stat.ethz.ch/R-manual/R-patched/library/base/html/file.path.html)

```{r}
for(i in 1:5){
  fileName = paste0("./data",i,".csv")
  print(fileName)
}
```




---

## Getting data off webpages 

<img class=center src=assets/img/googlescholar.png height='80%'/>

[http://scholar.google.com/citations?user=HI-I6C0AAAAJ&hl=en](http://scholar.google.com/citations?user=HI-I6C0AAAAJ&hl=en)

---

## Getting data off webpages


```{r}
library(XML)
con = url("http://scholar.google.com/citations?user=HI-I6C0AAAAJ&hl=en")
htmlCode = readLines(con)
close(con)
htmlCode
```



---

## Getting data off webpages

```{r}
html3 <- htmlTreeParse("http://scholar.google.com/citations?user=HI-I6C0AAAAJ&hl=en", useInternalNodes=T)
xpathSApply(html3, "//title", xmlValue)
xpathSApply(html3, "//td[@id='col-citedby']", xmlValue)
```





---

## Further resources

* Packages:
  * [httr](http://cran.r-project.org/web/packages/httr/index.html) - for working with http connections 
  * [RMySQL](http://cran.r-project.org/web/packages/RMySQL/index.html) - for interfacing with mySQL
  * [bigmemory](http://www.bigmemory.org/) - for handling data larger than RAM
  * [RHadoop](https://github.com/RevolutionAnalytics/RHadoop/wiki) - for interfacing R and Hadoop (by [Revolution Analytics](http://www.revolutionanalytics.com/))
  * [foreign](http://cran.r-project.org/web/packages/foreign/index.html) - for getting data into R from SAS, SPSS, Octave, etc. 
  
* Reading/writing R videos [Part 1](http://www.youtube.com/watch?v=aBzAels6jPk&list=PLjTlxb-wKvXNSDfcKPFH2gzHGyjpeCZmJ&index=8), [Part 2](http://www.youtube.com/watch?v=cUUqDWttMws&list=PLjTlxb-wKvXNSDfcKPFH2gzHGyjpeCZmJ&index=9)












