---
title       : Regression with factor variables
subtitle    : 
author      : Jeffrey Leek, Assistant Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow  # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## Key ideas

* Outcome is still quantitative
* Covariate(s) are factor variables
* Fitting lines = fitting means
* Want to evaluate contribution of all factor levels at once


---

## Example: Movie ratings

<img class=center src=assets/img/rotten.png height='80%'/>

[http://www.rottentomatoes.com/](http://www.rottentomatoes.com/)


---

## Movie Data

```{r loadMovies,cache=TRUE}
download.file("http://www.rossmanchance.com/iscam2/data/movies03RT.txt",destfile="./data/movies.txt")
movies <- read.table("./data/movies.txt",sep="\t",header=T,quote="")
head(movies)
```

[http://www.rossmanchance.com/](http://www.rossmanchance.com/)

---

## Rotton tomatoes score vs. rating

```{r, dependson="loadMovies",fig.height=4,fig.width=4}
plot(movies$score ~ jitter(as.numeric(movies$rating)),col="blue",xaxt="n",pch=19)
axis(side=1,at=unique(as.numeric(movies$rating)),labels=unique(movies$rating))
```


---

## Average score by rating

```{r, dependson="loadMovies",fig.height=4,fig.width=4}
plot(movies$score ~ jitter(as.numeric(movies$rating)),col="blue",xaxt="n",pch=19)
axis(side=1,at=unique(as.numeric(movies$rating)),labels=unique(movies$rating))
meanRatings <- tapply(movies$score,movies$rating,mean)
points(1:4,meanRatings,col="red",pch="-",cex=5)
```

---

## Another way to write it down

<br><br>

$$S_i = b_0 + b_1 \mathbb{1}(Ra_i="PG") + b_2 \mathbb{1}(Ra_i="PG-13") + b_3 \mathbb{1}(Ra_i="R") + e_i $$

<br><br>

The notation $\mathbb{1}(Ra_i="PG")$ is a logical value that is one if the movie rating is "PG" and zero otherwise.


__Average values__

$b_0$ = average of the G movies

$b_0 + b_1$ = average of the PG movies

$b_0 + b_2$ = average of the PG-13 movies

$b_0 + b_3$ = average of the R movies

---

## Here is how you do it in R

```{r lmFitChunk, dependson="loadMovies",fig.height=4,fig.width=4}
lm1 <- lm(movies$score ~ as.factor(movies$rating))
summary(lm1)
```


---

## Plot fitted values

```{r, dependson="lmFitChunk",fig.height=4,fig.width=4}
plot(movies$score ~ jitter(as.numeric(movies$rating)),col="blue",xaxt="n",pch=19)
axis(side=1,at=unique(as.numeric(movies$rating)),labels=unique(movies$rating))
points(1:4,lm1$coeff[1] + c(0,lm1$coeff[2:4]),col="red",pch="-",cex=5)
```



---

## Question 1

__Average values__

$b_0$ = average of the G movies

$b_0 + b_1$ = average of the PG movies

$b_0 + b_2$ = average of the PG-13 movies

$b_0 + b_3$ = average of the R movies


__What is the average difference in rating between G and R movies?__

$b_0 + b_3 - b_0$ = $b_3$

---

## Question 1 in R 

```{r, dependson="loadMovies",fig.height=4,fig.width=4}
lm1 <- lm(movies$score ~ as.factor(movies$rating))
summary(lm1)
```

---

## Question 1 in R 

```{r, dependson="loadMovies",fig.height=4,fig.width=4}
lm1 <- lm(movies$score ~ as.factor(movies$rating))
confint(lm1)
```


---

## Question 2

__Average values__

$b_0$ = average of the G movies

$b_0 + b_1$ = average of the PG movies

$b_0 + b_2$ = average of the PG-13 movies

$b_0 + b_3$ = average of the R movies


__What is the average difference in rating between $PG-13$ and $R$ movies?__

$b_0 + b_2 - (b_0 + b_3)$ = $b_2-b_3$


---

## We could rewrite our model


$$S_i = b_0 + b_1 \mathbb{1}(Ra_i="G") + b_2 \mathbb{1}(Ra_i="PG") + b_3 \mathbb{1}(Ra_i="PG-13") + e_i $$

<br><br>


__Average values__

$b_0$ = average of the R movies

$b_0 + b_1$ = average of the G movies

$b_0 + b_2$ = average of the PG movies

$b_0 + b_3$ = average of the PG-13 movies

__What is the average difference in rating between $PG-13$ and $R$ movies?__

$b_0 + b_3 - b_0$ = $b_3$


---

## Question 2 in R


```{r, dependson="lmFitChunk",fig.height=4,fig.width=4}
lm2 <- lm(movies$score ~ relevel(movies$rating,ref="R"))
summary(lm2)
```


---

## Question 2 in R


```{r, dependson="lmFitChunk",fig.height=4,fig.width=4}
lm2 <- lm(movies$score ~ relevel(movies$rating,ref="R"))
confint(lm2)
```

---

## Question 3


<br><br>

$$S_i = b_0 + b_1 \mathbb{1}(Ra_i="PG") + b_2 \mathbb{1}(Ra_i="PG-13") + b_3 \mathbb{1}(Ra_i="R") + e_i $$

<br><br>

__Average values__

$b_0$ = average of the G movies

$b_0 + b_1$ = average of the PG movies

$b_0 + b_2$ = average of the PG-13 movies

$b_0 + b_3$ = average of the R movies

__Is there any difference in score between any of the movie ratings?__

---

## Question 3 in R

```{r, dependson="loadMovies",fig.height=4,fig.width=4}
lm1 <- lm(movies$score ~ as.factor(movies$rating))
anova(lm1)
```


---

## Sum of squares (G movies)

```{r, dependson="lmFitChunk",fig.height=4,fig.width=4}
gMovies <- movies[movies$rating=="G",]; xVals <- seq(0.2,0.8,length=4)
plot(xVals,gMovies$score,ylab="Score",xaxt="n",xlim=c(0,1),pch=19)
abline(h=mean(gMovies$score),col="blue",lwd=3); abline(h=mean(movies$score),col="red",lwd=3)
segments(xVals+0.01,rep(mean(gMovies$score),length(xVals)),xVals+0.01,
         rep(mean(movies$score),length(xVals)),col="red",lwd=2)
segments(xVals-0.01,gMovies$score,xVals-0.01,rep(mean(gMovies$score),length(xVals)),col="blue",lwd=2)
```


---

## Tukey's (honestly significant difference test) 

```{r, dependson="loadMovies",fig.height=4,fig.width=4}
lm1 <- aov(movies$score ~ as.factor(movies$rating))
TukeyHSD(lm1)
```

[http://en.wikipedia.org/wiki/Tukey's_range_test](http://en.wikipedia.org/wiki/Tukey's_range_test)