---
title       : Getting Data (Part 1)
subtitle    : 
author      : Jeffrey Leek, Assistant Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn   # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 70)
opts_chunk$set(message = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache=F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```


## Get/set your working directory

Roger's lectures [windows](http://www.youtube.com/watch?v=XBcvH1BpIBo&list=PLjTlxb-wKvXNSDfcKPFH2gzHGyjpeCZmJ&index=2), [mac](http://www.youtube.com/watch?v=8xT3hmJQskU&list=PLjTlxb-wKvXNSDfcKPFH2gzHGyjpeCZmJ&index=1)
Andrew Jaffe's [lecture notes](https://dl.dropbox.com/u/7710864/courseraPublic/otherResources/lecture1/index.html)

```{r}
getwd()
setwd("/Users/jtleek/Dropbox/Jeff/teaching/2013/coursera/week2/004gettingData1/data")
getwd()
```

Important difference with Windows: 

```{r, eval=FALSE}
 setwd("C:\\Users\\Andrew\\Downloads")
```


---

## Get/set your working directory (relative paths)

```{r}
getwd()
setwd("./data")
getwd()
setwd("../")
getwd()
```


---

## Get/set your working directory (relative paths)

```{r}
getwd()
setwd("/Users/jtleek/Dropbox/Jeff/teaching/2013/coursera/week2/004gettingData1/data")
getwd()
```


---

## Types of files data may come from

* Tab-delimited text
* Comma-separated text
* Excel file
* JSON File 
* HTML/XML file
* Database


---

## Where you can get data

* From a colleague
* From the web
* From an application programming interface
* By scraping a web page 


---

## Getting data from the internet - download.file()

* Downloads a file from the internet
* Even if you could do this by hand, helps with reproducibility
* Important parameters are _url_, _destfile_, _method_
* Useful for downloading tab-delimited, csv, etc. 

---

## Example - Baltimore camera data

<img class=center src=assets/img/cameras.png height='80%'/>

[https://data.baltimorecity.gov/Transportation/Baltimore-Fixed-Speed-Cameras/dz54-2aru](https://data.baltimorecity.gov/Transportation/Baltimore-Fixed-Speed-Cameras/dz54-2aru)


---

## Example - Baltimore camera data,csv

<img class=center src=assets/img/cameraslink.png height='80%'/>

[https://data.baltimorecity.gov/Transportation/Baltimore-Fixed-Speed-Cameras/dz54-2aru](https://data.baltimorecity.gov/Transportation/Baltimore-Fixed-Speed-Cameras/dz54-2aru)


---

## Download a file from the web


```{r}
fileUrl <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.csv?accessType=DOWNLOAD"
download.file(fileUrl,destfile="./data/cameras.csv",method="curl")
list.files("./data")
dateDownloaded <- date()
dateDownloaded
```


---

## Some notes about download.file()

* If the url starts with _http_ you can use download.file()
* If the url starts with _https_ on Windows you may be ok
* If the url starts with _https_ on Mac you may need to set _method="curl"_
* If the file is big, this might take a while
* Be sure to record when you downloaded. 

---

## Loading data you have saved - read.table()

* This is the main function for reading data into R
* Flexible and robust but requires more parameters
* Reads the data into RAM - big data can cause problems
* Important parameters _file_, _header_, _sep_, _row.names_, _nrows_
* Related: _read.csv()_, _read.csv2()_

---

## Example: Baltimore camera data

```{r}
getwd()
cameraData <- read.table("./data/cameras.csv")
head(cameraData)

```


---

## Example: Baltimore camera data

```{r}
getwd()
cameraData <- read.table("./data/cameras.csv",sep=",",header=TRUE)
head(cameraData)

```


---

## Example: Baltimore camera data

read.csv sets _sep=","_ and _header=TRUE_ 
```{r}
cameraData <- read.csv("./data/cameras.csv")
head(cameraData)
```



---

## read.xlsx(), read.xlsx2() {xlsx package}


* Reads .xlsx files, but slow
* Important parameters _file_, _sheetIndex_, _sheetIndex_, _rowIndex_, _colIndex_, _header_
* read.xlsx2() relies more on low level Java functions so may be a bit faster


---

## read.xlsx() - Baltimore camera data

```{r}
library(xlsx)
fileUrl <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.xlsx?accessType=DOWNLOAD"
download.file(fileUrl,destfile="./data/camera.xlsx",method="curl")
cameraData <- read.xlsx2("./data/camera.xlsx",sheetIndex=1)
head(cameraData)
```


---

## Picking a file - less reproducible, but useful

```{r,eval=FALSE}
cameraData <- read.csv(file.choose())
```

<img class=center src=assets/img/filechoose.png height='80%'/>

