---
title       : Summarizing data
subtitle    : 
author      : Jeffrey Leek, Assistant Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn   # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 70)
opts_chunk$set(message = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache = T, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## Why summarize?

* Data are often too big to look at the whole thing
* The first step in an analysis is to find problems
* When you do these summaries you should be looking for
  * Missing values
  * Values outside of expected ranges
  * Values that seem to be in the wrong units
  * Mislabled variables/columns 
  * Variables that are the wrong class

---

## Earthquake data

<img class=center src=assets/img/earthquakes.png height='80%'/>

[https://explore.data.gov/Geography-and-Environment/Worldwide-M1-Earthquakes-Past-7-Days/7tag-iwnu](https://explore.data.gov/Geography-and-Environment/Worldwide-M1-Earthquakes-Past-7-Days/7tag-iwnu)


---

## Earthquake data

```{r cachedChunk,cache=TRUE}
fileUrl <- "http://earthquake.usgs.gov/earthquakes/catalogs/eqs7day-M1.txt"
download.file(fileUrl,destfile="./data/earthquakeData.csv",method="curl")
dateDownloaded <- date()
dateDownloaded
eData <- read.csv("./data/earthquakeData.csv")
```

---

## Looking at data - the whole thing
```{r,dependson="cachedChunk"}
eData
```

---

## Looking at data - dim(),names(),nrow(),ncol()
```{r, dependson="cachedChunk"}
dim(eData)
names(eData)
nrow(eData)
```

---

## Looking at the data - quantile(),summary()

```{r, dependson="cachedChunk"}
quantile(eData$Lat)
summary(eData)
```



---

## Looking at data - class()

```{r, dependson="cachedChunk"}
class(eData)
sapply(eData[1,],class)
```

---

## Looking at data - unique(),length(),table()

```{r, dependson="cachedChunk"}
unique(eData$Src)
length(unique(eData$Src))
table(eData$Src)
```

---

## Looking at data - table()

```{r, dependson="cachedChunk"}
table(eData$Src,eData$Version)
```


---

## Looking at data - any(), all()
```{r, dependson="cachedChunk"}
eData$Lat[1:10]
eData$Lat[1:10] > 40
any(eData$Lat[1:10] > 40)
```

---

## Looking at data - all()
```{r, dependson="cachedChunk"}

eData$Lat[1:10] > 40
all(eData$Lat[1:10] > 40)
```



---

## Looking at subsets - &

```{r, dependson="cachedChunk"}
eData[eData$Lat > 0 & eData$Lon > 0,c("Lat","Lon")]

```

---

## Looking at subsets - |

```{r, dependson="cachedChunk"}
eData[eData$Lat > 0 | eData$Lon > 0,c("Lat","Lon")]

```


---

## Peer review experiment data

* Data on submissions/reviews in an experiment 

<img class=center src=assets/img/cooperation.png height='60%'/>

[http://www.plosone.org/article/info:doi/10.1371/journal.pone.0026895](http://www.plosone.org/article/info:doi/10.1371/journal.pone.0026895)


---

## Peer review data


```{r reviewDownload, cache=TRUE}
fileUrl1 <- "https://dl.dropbox.com/u/7710864/data/reviews-apr29.csv"
fileUrl2 <- "https://dl.dropbox.com/u/7710864/data/solutions-apr29.csv"
download.file(fileUrl1,destfile="./data/reviews.csv",method="curl")
download.file(fileUrl2,destfile="./data/solutions.csv",method="curl")
reviews <- read.csv("./data/reviews.csv"); solutions <- read.csv("./data/solutions.csv")
head(reviews,2)
head(solutions,2)
```

---

## Find if there are missing values - is.na()

```{r, dependson="reviewDownload"}
is.na(reviews$time_left[1:10])
sum(is.na(reviews$time_left))
table(is.na(reviews$time_left))
```

---

## Important table()/NA issue

```{r}
table(c(0,1,2,3,NA,3,3,2,2,3))
table(c(0,1,2,3,NA,3,3,2,2,3),useNA="ifany")
```

---

## Summarizing columns/rows - rowSums(),rowMeans(),colSums(),colMeans()

* Important parameters: _x_, _na.rm_
```{r, dependson="reviewDownload"}
colSums(reviews)
```

---

## Summarizing columns/rows - rowSums(),rowMeans(),colSums(),colMeans()

```{r, dependson="reviewDownload"}
colMeans(reviews,na.rm=TRUE)
rowMeans(reviews,na.rm=TRUE)
```




