---
title       : Clustering example
subtitle    : 
author      : Jeffrey Leek, Assistant Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache = T, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## Samsung Galaxy S3

<img class=center src=assets/img/samsung.png height='80%'/>

[http://www.samsung.com/global/galaxys3/](http://www.samsung.com/global/galaxys3/)


---

## Samsung Data

<img class=center src=assets/img/ucisamsung.png height='60%'/>

[http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones)


---

## Slightly processed data

```{r loadData,cache=FALSE}
download.file("https://dl.dropbox.com/u/7710864/courseraPublic/samsungData.rda"
              ,destfile="./data/samsungData.rda",method="curl")
load("./data/samsungData.rda")
names(samsungData)[1:12]
table(samsungData$activity)
```

---

## Plotting average acceleration for first subject

```{r processData,dependson="loadData",fig.height=4.5,fig.width=8}
par(mfrow=c(1,2))
numericActivity <- as.numeric(as.factor(samsungData$activity))[samsungData$subject==1]
plot(samsungData[samsungData$subject==1,1],pch=19,col=numericActivity,ylab=names(samsungData)[1])
plot(samsungData[samsungData$subject==1,2],pch=19,col=numericActivity,ylab=names(samsungData)[2])
legend(150,-0.1,legend=unique(samsungData$activity),col=unique(numericActivity),pch=19)
```

---

## Clustering based just on average acceleration


```{r dependson="processData",fig.height=4,fig.width=4,cache=TRUE}
source("http://dl.dropbox.com/u/7710864/courseraPublic/myplclust.R")
distanceMatrix <- dist(samsungData[samsungData$subject==1,1:3])
hclustering <- hclust(distanceMatrix)
myplclust(hclustering,lab.col=numericActivity)
```


---

## Plotting max acceleration for the first subject

```{r ,dependson="processData",fig.height=4,fig.width=8}
par(mfrow=c(1,2))
plot(samsungData[samsungData$subject==1,10],pch=19,col=numericActivity,ylab=names(samsungData)[10])
plot(samsungData[samsungData$subject==1,11],pch=19,col=numericActivity,ylab=names(samsungData)[11])
```

---

## Clustering based on maximum acceleration

```{r dependson="processData",fig.height=4,fig.width=4,cache=TRUE}
source("http://dl.dropbox.com/u/7710864/courseraPublic/myplclust.R")
distanceMatrix <- dist(samsungData[samsungData$subject==1,10:12])
hclustering <- hclust(distanceMatrix)
myplclust(hclustering,lab.col=numericActivity)
```



---

## Singular value decomposition

```{r svdChunk,dependson="processData",fig.height=4,fig.width=8,cache=TRUE}
svd1 = svd(scale(samsungData[samsungData$subject==1,-c(562,563)]))
par(mfrow=c(1,2))
plot(svd1$u[,1],col=numericActivity,pch=19)
plot(svd1$u[,2],col=numericActivity,pch=19)
```

---

## Find maximum contributor

```{r dependson="svdChunk",fig.height=4,fig.width=4,cache=TRUE}
plot(svd1$v[,2],pch=19)

```


---

##  New clustering with maximum contributer

```{r dependson="svdChunk",fig.height=4.5,fig.width=4.5,cache=TRUE}
maxContrib <- which.max(svd1$v[,2])
distanceMatrix <- dist(samsungData[samsungData$subject==1,c(10:12,maxContrib)])
hclustering <- hclust(distanceMatrix)
myplclust(hclustering,lab.col=numericActivity)                             
```


---

##  New clustering with maximum contributer

```{r dependson="svdChunk",fig.height=4.5,fig.width=4.5,cache=TRUE}
names(samsungData)[maxContrib]                          
```

---

##  K-means clustering (nstart=1, first try)

```{r kmeans1,dependson="processData",fig.height=4,fig.width=4}
kClust <- kmeans(samsungData[samsungData$subject==1,-c(562,563)],centers=6)
table(kClust$cluster,samsungData$activity[samsungData$subject==1])
```



---

##  K-means clustering (nstart=1, second try)

```{r dependson="kmeans1",fig.height=4,fig.width=4,cache=TRUE}
kClust <- kmeans(samsungData[samsungData$subject==1,-c(562,563)],centers=6,nstart=1)
table(kClust$cluster,samsungData$activity[samsungData$subject==1])
```


---

##  K-means clustering (nstart=100, first try)

```{r dependson="kmeans1",fig.height=4,fig.width=4,cache=TRUE}
kClust <- kmeans(samsungData[samsungData$subject==1,-c(562,563)],centers=6,nstart=100)
table(kClust$cluster,samsungData$activity[samsungData$subject==1])
```



---

##  K-means clustering (nstart=100, second try)

```{r kmeans100,dependson="kmeans1",fig.height=4,fig.width=4,cache=TRUE}
kClust <- kmeans(samsungData[samsungData$subject==1,-c(562,563)],centers=6,nstart=100)
table(kClust$cluster,samsungData$activity[samsungData$subject==1])
```

---

##  Cluster 1 Variable Centers (Laying)

```{r dependson="kmeans100",fig.height=4,fig.width=8,cache=FALSE}
plot(kClust$center[1,1:10],pch=19,ylab="Cluster Center",xlab="")
```


---

##  Cluster 2 Variable Centers (Walking)

```{r dependson="kmeans100",fig.height=4,fig.width=8,cache=FALSE}
plot(kClust$center[6,1:10],pch=19,ylab="Cluster Center",xlab="")
```


