---
title       : Simulation basics
subtitle    : 
author      : Jeffrey Leek, Assistant Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn   # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---
```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 70)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache = T, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## Important simulation functions

__Distributions__
* rbeta, rbinom, rcauchy, rchisq, rexp, rf, rgamma, rgeom, rhyper, rlogis, rlnorm, rnbinom, rnorm, rpois, rt, runif, rweibull

__Densities__

* dbeta,dbinom, dcauchy, dchisq, dexp, df, dgamma, dgeom, dhyper, dlogis, dlnorm, dnbinom, dnorm, dpois, dt, dunif, dweibull

__Sampling__
* sample(,replace=TRUE),sample(replace=FALSE)


---

## r_foo_ functions generate data 

__Normal__
```{r}
args(rnorm)
heights = rnorm(10,mean=188,sd=3)
heights
```

---

## r_foo_ functions generate data

__Binomial__
```{r}
args(rbinom)
coinFlips = rbinom(10,size=10,prob=0.5)
coinFlips
```


---

## Example distribution: Normal

__Normal Distribution: $N(\mu,\sigma)$__
* $X \sim N(0,1)$

```{r, echo=FALSE, fig.height = 4, fig.width = 4}
xvals = seq(-10,10,length=100)
plot(xvals,dnorm(xvals,0,1),type="l",col="blue",lwd=3,ylab="density",xlab="X value")

```

---

## d_foo_ functions calculate the density
__Normal__
```{r}
args(dnorm)
x = seq(from=-5,to=5,length=10)
normalDensity = dnorm(x,mean=0,sd=1)
round(normalDensity,2)

```

---


## Example distribution: Binomial

__Binomial distribution: $Bin(n,p)$__
* $X \sim Bin(10,0.5)$
```{r, echo=FALSE, fig.height = 4, fig.width = 4}
xvals = 0:10
plot(xvals,dbinom(xvals,10,0.5),type="h",col="blue",lwd=3,ylab="probability",xlab="X value")

```

---
## d_foo_ functions calculate the density
__Binomial__
```{r}
args(dbinom)
x = seq(0,10,by=1)
binomialDensity = dbinom(x,size=10,prob=0.5)
round(binomialDensity,2)
```

---
## Sample draws a random sample
```{r}
args(sample)
heights = rnorm(10,mean=188,sd=3)
heights
sample(heights,size=10,replace=TRUE)

```

---

## Sample draws a random sample

```{r}
heights
sample(heights,size=10,replace=FALSE)

```

---

## Sample can draw according to a set of probabilities
```{r}
heights
probs = c(0.4,0.3,0.2,0.1,0,0,0,0,0,0)
sum(probs)
sample(heights,size=10,replace=TRUE,prob=probs)

```
---

## Setting a seed

Setting a seed ensures reproducible results from random processes in R
```{r}
set.seed(12345)
rnorm(5,mean=0,sd=1)

set.seed(12345)
rnorm(5,mean=0,sd=1)
```


---

## For more information

__More on distributions in R__

[http://cran.r-project.org/web/views/Distributions.html](http://cran.r-project.org/web/views/Distributions.html)

__Computing for Data Analysis__

[Simulation in R](http://www.youtube.com/watch?v=tvv4IA8PEzw&list=PLjTlxb-wKvXOzI2h0F2_rYZHIXz8GWBop&index=6)
