---
title       : Smoothing
subtitle    : 
author      : Jeffrey Leek, Assistant Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow  # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## Key ideas

* Sometimes there are non-linear trends in data
* We can use "smoothing" to try to capture these
* Still a risk of overfitting 
* Often hard to interpret

---

## CD4 Data

```{r cd4Data, cache=TRUE}
download.file("https://spark-public.s3.amazonaws.com/dataanalysis/cd4.data",
              destfile="./data/cd4.data",method="curl")
cd4Data <- read.table("./data/cd4.data", 
                      col.names=c("time", "cd4", "age", "packs", "drugs", "sex",
                                  "cesd", "id"))
cd4Data <- cd4Data[order(cd4Data$time),]
head(cd4Data)
```

[http://www.cbcb.umd.edu/~hcorrada/PracticalML/](http://www.cbcb.umd.edu/~hcorrada/PracticalML/)

---

## CD4 over time

```{r, dependson="cd4Data", cache=TRUE,fig.height=4,fig.width=4}
plot(cd4Data$time,cd4Data$cd4,pch=19,cex=0.1)
```

---

## Average first 2 points

```{r, dependson="cd4Data", cache=TRUE,fig.height=4,fig.width=4}
plot(cd4Data$time,cd4Data$cd4,pch=19,cex=0.1)
points(mean(cd4Data$time[1:2]),mean(cd4Data$cd4[1:2]),col="blue",pch=19)
```

---

## Average second and third points

```{r, dependson="cd4Data", cache=TRUE,fig.height=4,fig.width=4}
plot(cd4Data$time,cd4Data$cd4,pch=19,cex=0.1)
points(mean(cd4Data$time[1:2]),mean(cd4Data$cd4[1:2]),col="blue",pch=19)
points(mean(cd4Data$time[2:3]),mean(cd4Data$cd4[2:3]),col="blue",pch=19)
```


---

## A moving average

```{r, dependson="cd4Data", cache=TRUE,fig.height=4,fig.width=4}
plot(cd4Data$time,cd4Data$cd4,pch=19,cex=0.1)
aveTime <- aveCd4 <- rep(NA,length(3:(dim(cd4Data)[1]-2)))
for(i in 3:(dim(cd4Data)[1]-2)){
    aveTime[i] <- mean(cd4Data$time[(i-2):(i+2)])
    aveCd4[i] <- mean(cd4Data$cd4[(i-2):(i+2)])
}
lines(aveTime,aveCd4,col="blue",lwd=3)
```


---

## Average more points

```{r, dependson="cd4Data", cache=TRUE,fig.height=4,fig.width=4}
plot(cd4Data$time,cd4Data$cd4,pch=19,cex=0.1)
aveTime <- aveCd4 <- rep(NA,length(11:(dim(cd4Data)[1]-10)))
for(i in 11:(dim(cd4Data)[1]-2)){
  aveTime[i] <- mean(cd4Data$time[(i-10):(i+10)])
 aveCd4[i] <- mean(cd4Data$cd4[(i-10):(i+10)])
}
lines(aveTime,aveCd4,col="blue",lwd=3)
```

---

## Average many more

```{r, dependson="cd4Data", cache=TRUE,fig.height=4,fig.width=4}
plot(cd4Data$time,cd4Data$cd4,pch=19,cex=0.1)
aveTime <- aveCd4 <- rep(NA,length(201:(dim(cd4Data)[1]-200)))
for(i in 201:(dim(cd4Data)[1]-2)){
    aveTime[i] <- mean(cd4Data$time[(i-200):(i+200)])
    aveCd4[i] <- mean(cd4Data$cd4[(i-200):(i+200)])
}
lines(aveTime,aveCd4,col="blue",lwd=3)
```

---

## A faster way

```{r, dependson="cd4Data", cache=TRUE,fig.height=4,fig.width=4}
filtTime <- as.vector(filter(cd4Data$time,filter=rep(1,200))/200)
filtCd4 <- as.vector(filter(cd4Data$cd4,filter=rep(1,200))/200)
plot(cd4Data$time,cd4Data$cd4,pch=19,cex=0.1); lines(filtTime,filtCd4,col="blue",lwd=3)
```


---

## Averaging = weighted sums 

```{r, dependson="cd4Data", cache=TRUE,fig.height=4,fig.width=4}
filtCd4 <- as.vector(filter(cd4Data$cd4,filter=rep(1,4))/4)
filtCd4[2]
sum(cd4Data$cd4[1:4] * rep(1/4,4))
```


---

## Other weights -> should sum to one

```{r, dependson="cd4Data", cache=TRUE,fig.height=4,fig.width=4}
ws = 10; tukey = function(x) pmax(1 - x^2,0)^2
filt= tukey(seq(-ws,ws)/(ws+1));filt=filt/sum(filt)
plot(seq(-(ws),(ws)),filt,pch=19)
```

---

## Other weights -> should sum to one

```{r, dependson="cd4Data", cache=TRUE,fig.height=4,fig.width=4}
ws = 100; tukey = function(x) pmax(1 - x^2,0)^2
filt= tukey(seq(-ws,ws)/(ws+1));filt=filt/sum(filt)
filtTime <- as.vector(filter(cd4Data$time,filter=filt))
filtCd4 <- as.vector(filter(cd4Data$cd4,filter=filt))
plot(cd4Data$time,cd4Data$cd4,pch=19,cex=0.1); lines(filtTime,filtCd4,col="blue",lwd=3)
```

---

## Lowess (loess)

```{r lowess, dependson="cd4Data",fig.height=4,fig.width=4}
lw1 <- loess(cd4 ~ time,data=cd4Data)
plot(cd4Data$time,cd4Data$cd4,pch=19,cex=0.1)
lines(cd4Data$time,lw1$fitted,col="blue",lwd=3)
```


---

## Span

```{r, dependson="cd4Data",fig.height=4,fig.width=4}
plot(cd4Data$time,cd4Data$cd4,pch=19,cex=0.1,ylim=c(500,1500))
lines(cd4Data$time,loess(cd4 ~ time,data=cd4Data,span=0.1)$fitted,col="blue",lwd=3)
lines(cd4Data$time,loess(cd4 ~ time,data=cd4Data,span=0.25)$fitted,col="red",lwd=3)
lines(cd4Data$time,loess(cd4 ~ time,data=cd4Data,span=0.76)$fitted,col="green",lwd=3)
```


---

## Predicting with loess

```{r, dependson="lowess",fig.height=4,fig.width=4}
tme <- seq(-2,5,length=100); pred1 = predict(lw1,newdata=data.frame(time=tme),se=TRUE)
plot(tme,pred1$fit,col="blue",lwd=3,type="l",ylim=c(0,2500))
lines(tme,pred1$fit + 1.96*pred1$se.fit,col="red",lwd=3)
lines(tme,pred1$fit - 1.96*pred1$se.fit,col="red",lwd=3)
points(cd4Data$time,cd4Data$cd4,pch=19,cex=0.1)
```


---

## Splines

$$ Y_i = b_0 + \sum_{k=1}^K b_k s_k(x_i) + e_i $$


$Y_i$ - outcome for $i$th observation

$b_0$ - Intercept term

$b_k$ - Coefficient for $k$th spline function

$s_k$ - $k$th spline function

$x_i$ - covariate for $i$th observation

$e_i$ - everything we didn't measure/model


---

## Splines in R 

```{r splines, dependson="cd4Data",fig.height=4,fig.width=8}
library(splines)
ns1 <- ns(cd4Data$time,df=3)
par(mfrow=c(1,3))
plot(cd4Data$time,ns1[,1]); plot(cd4Data$time,ns1[,2]); plot(cd4Data$time,ns1[,3])
```

---

## Regression with splines

```{r splineReg, dependson="splines",fig.height=4,fig.width=4}
lm1 <- lm(cd4Data$cd4 ~ ns1)
summary(lm1)
```

---

## Fitted values

```{r, dependson="splineReg",fig.height=4,fig.width=4}
plot(cd4Data$time,cd4Data$cd4,pch=19,cex=0.1)
points(cd4Data$time,lm1$fitted,col="blue",pch=19,cex=0.5)
```

---

## Notes and further resources

__Notes__:

* Cross-validation with splines/smoothing is a good idea
* Do not predict outside the range of observed data

__Further resources__:

* [Hector Corrada Bravo's Lecture Notes](http://www.cbcb.umd.edu/~hcorrada/PracticalML/pdf/lectures/smoothing.pdf)
* [Rafa Irizarry's Lecture Notes on smoothing](http://www.biostat.jhsph.edu/~ririzarr/Teaching/649/section-06.pdf), [On splines](http://www.biostat.jhsph.edu/~ririzarr/Teaching/649/section-09.pdf)
* [Elements of Statistical Learning](http://www-stat.stanford.edu/~tibs/ElemStatLearn/)
* [Advanced Data Analysis from An Elementary Point of View](http://www.stat.cmu.edu/~cshalizi/ADAfaEPoV/ADAfaEPoV.pdf)



