---
title       : Predicting with regression models
subtitle    : 
author      : Jeffrey Leek, Assistant Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow  # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## Key ideas

* Use a standard regression model
  * lm
  * glm
* Predict new values with the coefficients
* Useful when the linear model is (nearly) correct

__Pros__:
* Easy to implement
* Easy to interpret

__Cons__:
* Often poor performance in nonlinear settings


---

## Example: Old faithful eruptions

<img class=center src=assets/img/yellowstone.png height='60%'/>

Image Credit/Copyright Wally Pacholka [http://www.astropics.com/](http://www.astropics.com/)

---

## Example: Old faithful eruptions

```{r faith}
data(faithful)
dim(faithful)
set.seed(333)
trainSamples <- sample(1:272,size=(272/2),replace=F)
trainFaith <- faithful[trainSamples,]
testFaith <- faithful[-trainSamples,]
head(trainFaith)
```

---

## Eruption duration versus waiting time

```{r dependson="faith",fig.height=4,fig.width=4}
plot(trainFaith$waiting,trainFaith$eruptions,pch=19,col="blue",xlab="Waiting",ylab="Duration")
```

---

## Fit a linear model 

$$ ED_i = b_0 + b_1 WT_i + e_i $$

```{r faithlm,dependson="faith",fig.height=4,fig.width=4}
lm1 <- lm(eruptions ~ waiting,data=trainFaith)
summary(lm1)
```


---
## Model fit

```{r dependson="faithlm",fig.height=4,fig.width=4}
plot(trainFaith$waiting,trainFaith$eruptions,pch=19,col="blue",xlab="Waiting",ylab="Duration")
lines(trainFaith$waiting,lm1$fitted,lwd=3)
```

---

## Predict a new value

$$\hat{ED} = \hat{b}_0 + \hat{b}_1 WT$$

```{r ,dependson="faithlm",fig.height=4,fig.width=4}
coef(lm1)[1] + coef(lm1)[2]*80
newdata <- data.frame(waiting=80)
predict(lm1,newdata)
```

---

## Plot predictions - training and test

```{r ,dependson="faithlm",fig.height=4,fig.width=8}
par(mfrow=c(1,2))
plot(trainFaith$waiting,trainFaith$eruptions,pch=19,col="blue",xlab="Waiting",ylab="Duration")
lines(trainFaith$waiting,predict(lm1),lwd=3)
plot(testFaith$waiting,testFaith$eruptions,pch=19,col="blue",xlab="Waiting",ylab="Duration")
lines(testFaith$waiting,predict(lm1,newdata=testFaith),lwd=3)
```

---

## Get training set/test set errors

```{r ,dependson="faithlm",fig.height=4,fig.width=4}
# Calculate RMSE on training
sqrt(sum((lm1$fitted-trainFaith$eruptions)^2))

# Calculate RMSE on test
sqrt(sum((predict(lm1,newdata=testFaith)-testFaith$eruptions)^2))
```

---

## Prediction intervals

```{r ,dependson="faithlm",fig.height=4,fig.width=4}
pred1 <- predict(lm1,newdata=testFaith,interval="prediction")
ord <- order(testFaith$waiting)
plot(testFaith$waiting,testFaith$eruptions,pch=19,col="blue")
matlines(testFaith$waiting[ord],pred1[ord,],type="l",,col=c(1,2,2),lty = c(1,1,1), lwd=3)
```

---

## Example with binary data: Baltimore Ravens

<img class=center src=assets/img/ravens.png height='60%'/>

[http://espn.go.com/nfl/team/_/name/bal/baltimore-ravens](http://espn.go.com/nfl/team/_/name/bal/baltimore-ravens)

---

## Ravens Data

```{r loadRavens,cache=TRUE}
download.file("https://dl.dropbox.com/u/7710864/data/ravensData.rda",
              destfile="./data/ravensData.rda",method="curl")
load("./data/ravensData.rda")
head(ravensData)
```

---

## Fit a logistic regression

$$logit(E[RW_i | RS_i]) = b_0 + b_1 RS_i$$
```{r glmRavens, dependson="loadRavens",fig.height=4,fig.width=8}
glm1 <- glm(ravenWinNum ~ ravenScore,family="binomial",data=ravensData)
par(mfrow=c(1,2))
boxplot(predict(glm1) ~ ravensData$ravenWinNum,col="blue")
boxplot(predict(glm1,type="response") ~ ravensData$ravenWinNum,col="blue")
```

---

## Choosing a cutoff (re-substitution)

```{r cutoff, dependson="glmRavens",fig.height=4,fig.width=4}
xx <- seq(0,1,length=10); err <- rep(NA,10)
for(i in 1:length(xx)){
  err[i] <- sum((predict(glm1,type="response") > xx[i]) != ravensData$ravenWinNum)
}
plot(xx,err,pch=19,xlab="Cutoff",ylab="Error")
```

---

## Comparing models with cross validation

```{r,dependson="glmRavens",fig.height=4,fig.width=4}
library(boot)
cost <- function(win, pred = 0) mean(abs(win-pred) > 0.5)
glm1 <- glm(ravenWinNum ~ ravenScore,family="binomial",data=ravensData)
glm2 <- glm(ravenWinNum ~ ravenScore,family="gaussian",data=ravensData)
cv1 <- cv.glm(ravensData,glm1,cost,K=3)
cv2 <- cv.glm(ravensData,glm2,cost,K=3)
cv1$delta
cv2$delta
```

---

## Notes and further reading

* Regression models with multiple covariates can be included
* Often useful in combination with other models 
* [Elements of statistical learning](http://www-stat.stanford.edu/~tibs/ElemStatLearn/)
* [Modern applied statistics with S](http://www.amazon.com/Modern-Applied-Statistics-W-N-Venables/dp/0387954570)
