---
title       : ANOVA with multiple factors/variables
subtitle    : 
author      : Jeffrey Leek, Assistant Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow  # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## Key ideas

* Outcome is still quantitative
* You have multiple explanatory variables
* Goal is to identify contributions of different variables

---

## A successful example

<img class=center src=assets/img/wired.png height='40%'/>

"For the button, an A/B test of three new word choices—”Learn More,” “Join Us Now,” and “Sign Up Now”—revealed that “Learn More” garnered 18.6 percent more signups per visitor than the default of “Sign Up.” Similarly, a black-and-white photo of the Obama family outperformed the default turquoise image by 13.1 percent. Using both the family image and “Learn More,” signups increased by a thundering 40 percent."

[http://www.wired.com/business/2012/04/ff_abtesting/](http://www.wired.com/business/2012/04/ff_abtesting/)

---

## Movie Data

```{r loadMovies,cache=FALSE}
download.file("http://www.rossmanchance.com/iscam2/data/movies03RT.txt",
              destfile="./data/movies.txt")
movies <- read.table("./data/movies.txt",sep="\t",header=T,quote="")
head(movies)
```

[http://www.rossmanchance.com/](http://www.rossmanchance.com/)


---

## Relating score to rating

<br><br>

$$S_i = b_0 + b_1 \mathbb{1}(Ra_i="PG") + b_2 \mathbb{1}(Ra_i="PG-13") + b_3 \mathbb{1}(Ra_i="R") + e_i $$

<br><br>

The notation $\mathbb{1}(Ra_i="PG")$ is a logical value that is one if the movie rating is "PG" and zero otherwise.


__Average values__

$b_0$ = average of the G movies

$b_0 + b_1$ = average of the PG movies

$b_0 + b_2$ = average of the PG-13 movies

$b_0 + b_3$ = average of the R movies

---

## ANOVA in R


```{r anovaMovies, dependson="loadMovies",cache=FALSE}
aovObject <- aov(movies$score ~ movies$rating)
aovObject
```

---

## ANOVA in R

```{r, dependson="anovaMovies"}
aovObject$coeff
```



---

## Adding a second factor

<br><br>

$$S_i = b_0 + b_1 \mathbb{1}(Ra_i="PG") + b_2 \mathbb{1}(Ra_i="PG-13") + b_3 \mathbb{1}(Ra_i="R")$$
$$+ \gamma_1 \mathbb{1}(G_i="action") +  \gamma_2 \mathbb{1}(G_i="animated") + ... + e_i$$

<br><br>

The notation $\mathbb{1}(Ra_i="PG")$ is a logical value that is one if the movie rating is "PG" and zero otherwise. 



---

## Adding a second factor

<br><br>

$$S_i = b_0 + \underbrace{b_1 \mathbb{1}(Ra_i="PG") + b_2 \mathbb{1}(Ra_i="PG-13") + b_3 \mathbb{1}(Ra_i="R")}_{rating}$$
$$+ \gamma_1 \underbrace{\mathbb{1}(G_i="action")  + \gamma_2 \mathbb{1}(G_i="animated") + ...}_{genre} + e_i $$

<br><br>

There are only 2 variables in this model. They have multiple levels. 


---

## Second variable


```{r twoVariableANOVA, dependson="loadMovies"}
aovObject2 <- aov(movies$score ~ movies$rating + movies$genre)
aovObject2
```

---

## ANOVA Summary


```{r, dependson="twoVariableANOVA"}
summary(aovObject2)
```

---

## Order matters

```{r, dependson="loadMovies"}
aovObject3 <- aov(movies$score ~ movies$genre + movies$rating)
summary(aovObject3)
summary(aovObject2)
```

---

## Adding a quantitative variable

<br><br>

$$S_i = b_0 + \underbrace{b_1 \mathbb{1}(Ra_i="PG") + b_2 \mathbb{1}(Ra_i="PG-13") + b_3 \mathbb{1}(Ra_i="R")}_{rating}$$
$$+ \gamma_1 \underbrace{\mathbb{1}(G_i="action")  + \gamma_2 \mathbb{1}(G_i="animated") + ...}_{genre} + \eta_1 \underbrace{BO_i}_{box\; office} + e_i $$

<br><br>

There are three variables in this model - box office is quantitative so only has one term. 

---

## ANOVA with quantitative variable in R

```{r, dependson="loadMovies"}
aovObject4 <- aov(movies$score ~ movies$genre + movies$rating + movies$box.office)
summary(aovObject4)
```


---

## Language and further resources

* Units - one observation
* Treatments - applied to units
* Factors - controlled by experimenters
* Replicates - multiple (independent) units with the same factors/treatments

* [Wikipedia on Experimental Design](http://en.wikipedia.org/wiki/Design_of_experiments)
* [Wikipedia on ANOVA](http://en.wikipedia.org/wiki/Analysis_of_variance)
* [Wikipedia on A/B Testing](http://en.wikipedia.org/wiki/A/B_testing)









